\titletext{}

### 2021-01-25 manuscript/regional_befr_manuscript.tex
Added submission

@@

% Custom title formatting
\let\oldtitle\title

\renewcommand{\title}[1]{\oldtitle{\vspace{-1.5cm}#1}}

\usepackage[breaklinks]{hyperref}
\definecolor{links}{RGB}{191,59,72}
\hypersetup{
	breaklinks,
	colorlinks,
	allcolors=links,
	linktoc=section,
	pdfauthor={John L. Godlee}
}
\def\subsectionautorefname{section}
\def\subsubsectionautorefname{section}
% \usepackage{biblatex}

@@

\usepackage{lineno}
\linenumbers

\newcommand{\titletext}{Stem density drives biodiversity-ecosystem function across disturbed southern African woodlands}

\title{\titletext{}}
\author{}
\date{}

@@

{\LARGE{\textbf{Blinded Main Text File}}}

\LARGE{Title: \titletext{}}

\normalsize{Running title: Diversity - ecosystem function in southern African woodlands}

\section*{Abstract}

\begin{itemize}
	\item{Positive biodiversity-ecosystem function relationships (BEFRs) have been widely documented, but whether BEFRs should be expected in disturbance-driven systems is less clear. Disturbance may limit competition and niche differentiation, which are frequently posited to underlie BEFRs. We explore the relationship between tree species diversity and biomass, one measure of ecosystem function, across southern African woodlands and savannas, an ecological system rife with disturbance from fire, herbivores and humans.}
	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability, disturbance by fire, stem density and vegetation type.}
	\item{We found positive effects of tree species diversity on aboveground biomass, operating via increased structural diversity. The observed BEFR was highly dependent on stem density, with a minimum threshold of \textapprox{}180 stems ha\textsuperscript{-1}. We also found that resource availability mainly affects biomass indirectly, via increasing species diversity.}
	\item{The study underlines the close association between tree diversity, ecosystem structure, environment and function in highly disturbed southern African savannas and woodlands. We suggest that tree diversity is an under-appreciated determinant of savanna and woodland structure and function.}
\end{itemize}

\textbf{Keywords:} biodiversity, ecosystem function, woodland, miombo, biomass, structural equation modelling, forest structure.

@@

% Ecosystem function definition and mechanisms
Ecosystem functions can be defined in broad terms as rate processes and aggregate properties of ecosystems that describe the nature of biotic activity within those ecosystems \citep{Jax2005}. This includes processes such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resilience of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass. The frequently reported BEF relationship invokes three main mechanisms to explain it \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater biodiversity fill a greater breadth of realised niche space and avoid competition due to differences in their resource acquisition strategies; 2) selection effects, whereby communities with greater biodiversity are more likely to include a species that contributes highly to the measured ecosystem function; and 3) facilitation effects, whereby communities with greater biodiversity are more likely to include combinations of species which together increase the others' functional contribution.

% Savannas are big and ecologically important
Savannas and woodlands are the dominant vegetation type across southern Africa, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}). The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin, and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}. Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages. These retain the common features of an open tree canopy and an understorey generally dominated by C4 grasses. Southern African savannas and woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which >300 are trees \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under international efforts to reduced deforestation and degradation (REDD+) \citep{Hinsley2015}. Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely unquantified consequences for ecosystem structure and function.

% Savannas might not be the same as forests
Compared to forest ecosystems, southern African dry tropical woodlands and savannas are highly structured by disturbance, through fire \citep{Lehmann2014}, herbivory \citep{Sankaran2008, Levick2009}, and human activities such as shifting cultivation agriculture \citep{Heinimann2017}, timber extraction and charcoal processing \citep{Dewees2010, McNicol2018b}. High levels of disturbance, by fire or otherwise, may weaken the role of competition in determining local species distribution. Disturbance reduces stem density and woody biomass, reducing competitive interactions between individuals, allowing weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}. This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as important in dry woodland/savanna ecosystems, thus weakening the BEF relationship. Instead, stress tolerance and the functional contribution of particular species (selection effects) may be the predominant forces influencing ecosystem function \citep{Lasky2014, Tobner2016}. A threshold stem density may exist below which the effects of tree species diversity on ecosystem function are not detectable, with potential consequences for our classification of ecosystems limited by biodiversity and those limited by other factors.

% Savannas might no be the same as forests continued: facilitation effects
More diverse species assemblages may lead to facilitation effects between certain species combinations under the limiting environmental conditions prevalent across African savannas, such as low water availability. Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments. They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, as competition diminishes in ecosystems where environmental stress limits individual species' abundances, thus reducing the competition which drives niche complementarity effects. This potential mismatch in the contribution of different mechanisms to the BEF relationship between dry tropical woodlands and other forested ecosystems demands further investigation if we are to derive a generalisable BEF relationship.

% African savannas are understudied in this regard
The representation of dry tropical ecosystems in the BEF relationship literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEF relationship studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, a narrow geographic region \citep{Barthlott2005}. A severe lack of study in dry tropical ecosystems, especially given the potential divergence in BEF relationship mechanisms described above, suggests that a focus on the BEF in southern African woodlands could greatly strengthen our understanding of a global BEF relationship and its environmental determinants. A small number of studies in southern African woodlands, all of which were restricted in spatial scope to a small region of miombo woodland, found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. The results of these fine-scale studies concur with similar studies in other biomes \citep{Cardinale2009}. Studies of the BEF relationship often find that at fine spatial scales (<1 ha), biodiversity shows a strong effect on ecosystem function, but at broad spatial scales  (>10000s ha) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.

@@

% Disturbance deserves another look
High levels of disturbance in southern African woodlands may moderate the observable BEF relationship through its effect on ecosystem composition. Fire disturbance in forests has been linked to abundance-dependent mortality among smaller trees \citep{Roques2001, Staver2009, Bond2005}. Some species in the regional species pool may be excluded from woodland plots with high levels of disturbance if they are unable to escape the fire bottleneck and grow to become a large tree. Selection effects may therefore be more important in maximising ecosystem function in disturbance-prone woodlands. If a given woodland plant community contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high biomass under an intense disturbance regime.

@@

\section{Materials and methods}

@@

% Situation
The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}; \citealp{White1987}). The study area spans the core climate space of the region, with a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. A 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \citep{White1987}, using WorldClim estimates of Mean Annual Temperature (MAT, BIO1) and Mean Annual Precipitation (MAP, BIO12) between 1970 and 2000 with a pixel size of 30 arc seconds (926 m at equator) \citep{Fick2017}.

% Plots chosen
Plots were chosen from a larger pool of 5395 plots held in the SEOSAW database \citep{seosaw_web} based on the quality and completeness of data collection, and plot setup. Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool. Only plots >0.1 hectares were used in analyses, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates. Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than ``grassy savanna'', which is considered a separate biome with very different species composition \citep{Parr2014}.

@@

% WorldClim, SoilGrids and MODIS
Climatic data were taken from the WorldClim database, using the BioClim variables \citep{Fick2017}. In addition to MAT and MAP, temperature stress was calculated as the mean diurnal temperature range (BIO2) and precipitation seasonality was calculated as the mean of the coefficient of variation of monthly mean precipitation (BIO15). Soil fertility data were extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot centre \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC) (cmolc kg\textsuperscript{-1}), soil organic carbon stocks (kg m\textsuperscript{-2}) percentage soil sand content (0.05-2 mm) by weight and soil nitrogen content (g kg\textsuperscript{-1}). These data are a modelled product derived from various remotely sensed and directly measured data sources. The degree of fire disturbance was calculated using the MODIS monthly burned area product at 500 m resolution (MCD64A1; \citealt{MODIS_burn}), counting the total number of times the plot pixel was classified as burning, between 2001 and 2018. We initially aimed to include disturbance by herbivory in our model, including total herbivore biomass from the \citet{Hempson2017} modelled herbivory product, but this inclusion prevented models from converging due to its collinearity with other observed variables, notably MAP and disturbance by fire.

@@

Estimated tree species richness was calculated for each plot using `ChaoRichness()' from the `iNEXT' package in R \citep{Hsieh2016}. This procedure extrapolates a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value. Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort among plots. Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}. To measure tree species evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} was calculated as the ratio of the estimated Shannon diversity index to the natural log of estimated species richness. Abundance evenness allows for greater niche complementarity at small scales due to potentially increased heterogeneity of functional traits. We quantified tree structural diversity for each plot by calculating the coefficient of variation of DBH (DBH CoV) and tree height (Height CoV).

@@

We used Structural Equation Modelling (SEM) to investigate the determinants of AGB. All SEMs were constructed and analysed in the `lavaan' package \citep{lavaan} in R version 3.6.0 \citep{R2019}. SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}. A key aspect to our decision to use SEM is that they can explicitly model and partition variance attributed to indirect effects, which is challenging in standard multiple regressions. Using SEMs also allowed us to describe latent variables such as ``water availability'', ``soil fertility'', and ``disturbance'' which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having directly observable measures in our dataset. SEM is also necessary to properly account for potential feedback mechanisms between aspects of environment and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to the covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.

Prior to analysis, we specified a conceptual model with factors expected to affect AGB: water availability, soil fertility, disturbance, tree species diversity, tree structural diversity and stem density (\autoref{con_mod}).

Observed variables were transformed to achieve normality where necessary and standardised to Z-scores prior to analysis (\hyperref[appendixa]{Appendix A}). Standardisation allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect size, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{Beaujean2014}. Standardisation also controls for variables with variation across different orders of magnitude, which could otherwise prevent adequate model estimation from the covariance matrix in `lavaan'. To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables had their sign reversed. For example, overall water availability is expected to decrease as soil sand content increases, therefore sand content was reversed for use in the water availability latent variable. Precipitation seasonality, and temperature stress were also reversed in this way to account for the direction of their effect on water availability.

@@

We assessed the role of tree species diversity and tree structural diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity. Structural diversity can also directly influence AGB in this model, separate to the effect of of species diversity. To account for variation in stem density, which may covary with species diversity, we included it as an observed variable in our model. To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation type separately. We compared unstandardised path coefficients among the models for different vegetation types to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stem density and AGB. Path coefficients show the effect of a given path with other paths held constant. Models were estimated using the ``MLM'' estimator, because it is robust to multivariate non-normality \citep{Shapiro1983}. Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error of Approximation (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB. We critically assessed model fit in each case, taking into consideration the recommendations of \citet{Hu1999} who define threshold values of acceptability for these model fit indices: CFI >0.85, TLI >0.85, RMSEA <0.15, alongside our judgement of the model estimates.

@@

% Correlations
Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity and structural diversity variables had moderate positive correlations with AGB. Stem density had the strongest correlation with AGB of all variables considered (\ccib{}). Environmental variables had weaker correlations with AGB than diversity variables, with all environmental variables having significant correlations with AGB, except fire frequency. The direction of these correlations was used as a test of our assumptions for the direction of influence of latent variables later used in the SEMs. MAP had positive correlations with all tree species diversity and structural diversity variables. Tree species diversity variables had clear positive correlations with stem density (species richness: \ccsi{}; Shannon equitability: \ccei{}), but structural diversity variables showed weak correlations with stem density (DBH CoV: \ccdvi{}, Height CoV: \cchvi{}).

@@

In our sub-sampling of the plot dataset by stem density, we found an increasing positive effect of tree species diversity on AGB as stem density increased (\autoref{sem_struc_stems_ha}). There appears to be a minimum stem density threshold at \textapprox{}180 trees ha\textsuperscript{-1} below which there appears to be a reasonably constant baseline effect of tree diversity on biomass. The effect of structural diversity on AGB appears to remain constant with increasing stem density. The indirect effect of tree species diversity on AGB via structural diversity climbs as stem density increases.

@@

A model incorporating the latent variables of water availability, soil fertility and disturbance by fire showed that the total effect of tree species diversity on biomass was similar to that of water availability, soil fertility and disturbance (\autoref{full_mod}, \hyperref[appendixd]{Appendix D}). The direct effects of water availability, soil fertility and disturbance on AGB were negligible (water: \fmbetamb{}, soil: \fmbetasb{}, disturbance: \fmbetafb{}), with nearly all of their observed effect on AGB coming from the indirect paths via stem density (water: \fmbetamib{}, soil: \fmbetasib{}, disturbance: \fmbetafib{}) and species diversity (water: \fmbetamd{}, soil: \fmbetasd{}, disturbance: \fmbetafd{}). MAP and soil sand content had the greatest contributions to the latent variable of water availability. Model fit was acceptable: CFI = 0.925, TLI = 0.900, and RMSEA = \fmrmsea{}, R\textsuperscript{2} of AGB = \fmrsq{}.

@@

We found a non-linear positive effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}). At low stem densities, competition between mature trees may not occur, meaning that the niche complementarity effect provided by an increase in tree species richness may not be present, accounting for the small effect of tree species diversity on AGB below \textapprox{}180 trees ha\textsuperscript{-1}. At very high stem density, there is also an increase in the effect of species diversity on structural diversity. This could be because at high stem density, the adaptation of different species to growth form become important. At low stem density, individual trees tend to spread out rather than growing tall, but at high stem density, only certain species are able to exist in the understorey, while others are able to grow tall above the woodland canopy, leading to greater variation in tree height over the plot.

@@

\subsection{Conclusion}

@@

\section{Author contribution statement}

>>>>

\usepackage{framed}

@@

\usepackage[breaklinks]{hyperref}
\definecolor{links}{RGB}{0,0,0}
\hypersetup{
	breaklinks,
	colorlinks=true,
	linkcolor=links,
	anchorcolor=links,
	citecolor=links,
	filecolor=links,
	menucolor=links,
	runcolor=links,
	urlcolor=links,
	pdfauthor={John L. Godlee}
}
\def\subsectionautorefname{section}
\def\subsubsectionautorefname{section}
% \usepackage{biblatex}

@@

\linenumbers
\linespread{1.25}

\newcommand{\titletext}{Structural diversity and tree density drives variation in the biodiversity-ecosystem function relationship of woodlands and savannas}

@@

{\Large{Title: \titletext{}}}

\section*{Summary}

\begin{itemize}
	\item{Positive biodiversity-ecosystem function relationships (BEFRs) have been widely documented, but it is unclear if BEFRs should be expected in disturbance-driven systems. Disturbance may limit competition and niche differentiation, which are frequently posited to underlie BEFRs. We explore the relationship between tree species diversity and biomass, one measure of ecosystem function, across southern African woodlands and savannas, an ecological system rife with disturbance from fire, herbivores and humans.}
	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, accounting for interactive effects of resource availability, disturbance by fire, stem density and vegetation type.}
	\item{We found positive effects of tree species diversity on aboveground biomass, operating via increased structural diversity. The observed BEFR was highly dependent on stem density, with a minimum threshold of c. 180 stems ha\textsuperscript{-1}. We found that resource availability mainly affects biomass indirectly, via increasing species diversity.}
	\item{The study underlines the close association between tree diversity, ecosystem structure, environment and function in highly disturbed savannas and woodlands. We suggest that tree diversity is an under-appreciated determinant of savanna and woodland structure and function.}
\end{itemize}

\textbf{Keywords:} biodiversity, biomass, ecosystem function, forest structure, miombo, savanna, structural equation modelling, woodland.

@@

% Ecosystem function definition and mechanisms
Ecosystem functions can be defined in broad terms as rate processes and aggregate properties of ecosystems that describe the nature of biotic activity within those ecosystems \citep{Jax2005}. This includes processes such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resilience of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass. The frequently reported BEFR invokes three main mechanisms to explain it \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater biodiversity fill a greater breadth of realised niche space and avoid competition due to differences in their resource acquisition strategies; 2) selection effects, whereby communities with greater biodiversity are more likely to include a species that contributes highly to the measured ecosystem function; and 3) facilitation effects, whereby communities with greater biodiversity are more likely to include combinations of species which together increase the others' functional contribution.

% Savannas are big and ecologically important
Savannas and woodlands are the dominant vegetation type across southern Africa, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}). The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin, and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}. Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages. These retain the common features of an open tree canopy and an understorey generally dominated by C4 grasses. Southern African savannas and woodlands are highly diverse, thought to harbour c. 8500 plant species of which >300 are trees \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under international efforts to reduced deforestation and degradation (REDD+) \citep{Hinsley2015}. Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely unquantified consequences for ecosystem structure and function.

% Savannas might not be the same as forests
Compared to forest ecosystems, southern African dry tropical woodlands and savannas are highly structured by disturbance, through fire \citep{Lehmann2014}, herbivory \citep{Sankaran2008, Levick2009}, and human activities such as shifting cultivation agriculture \citep{Heinimann2017}, timber extraction and charcoal processing \citep{Dewees2010, McNicol2018b}. High levels of disturbance, by fire or otherwise, may weaken the role of competition in determining local species distribution. Disturbance reduces stem density and woody biomass, reducing competitive interactions between individuals, allowing weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}. This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as important in dry woodland/savanna ecosystems, thus weakening the BEFR. Instead, stress tolerance and the functional contribution of particular species (selection effects) may be the predominant forces influencing ecosystem function \citep{Lasky2014, Tobner2016}. A threshold stem density may exist below which the effects of tree species diversity on ecosystem function are not detectable, with potential consequences for our classification of ecosystems limited by biodiversity and those limited by other factors.

% Savannas might no be the same as forests continued: facilitation effects
More diverse species assemblages may lead to facilitation effects between certain species combinations under the limiting environmental conditions prevalent across African savannas, such as low water availability. Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments. They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, as competition diminishes in ecosystems where environmental stress limits individual species' abundances, thus reducing the competition which drives niche complementarity effects. This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation if we are to derive a generalisable BEFR.

% African savannas are understudied in this regard
The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, a narrow geographic region \citep{Barthlott2005}. A severe lack of study in dry tropical ecosystems, especially given the potential divergence in BEFR mechanisms described above, suggests that a focus on the BEFR in southern African woodlands could greatly strengthen our understanding of a global BEFR and its environmental determinants. A small number of studies in southern African woodlands, all of which were restricted in spatial scope to a small region of miombo woodland, found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. The results of these fine-scale studies concur with similar studies in other biomes \citep{Cardinale2009}. Studies of the BEFR often find that at fine spatial scales (<1 ha), biodiversity shows a strong effect on ecosystem function, but at broad spatial scales  (>10000s ha) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.

@@

% Disturbance deserves another look
High levels of disturbance in southern African woodlands may moderate the observable BEFR through its effect on ecosystem composition. Fire disturbance in forests has been linked to abundance-dependent mortality among smaller trees \citep{Roques2001, Staver2009, Bond2005}. Some species in the regional species pool may be excluded from woodland plots with high levels of disturbance if they are unable to escape the fire bottleneck and grow to become a large tree. Selection effects may therefore be more important in maximising ecosystem function in disturbance-prone woodlands. If a given woodland plant community contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high biomass under an intense disturbance regime.

@@

\section{Materials and Methods}

@@

% Situation
The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}, \citealp{White1987}). The study area spans the core climate space of the region, with a precipitation gradient from c. 460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to c. 1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. A 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \citep{White1987}, using WorldClim estimates of Mean Annual Temperature (MAT, BIO1) and Mean Annual Precipitation (MAP, BIO12) between 1970 and 2000 with a pixel size of 30 arc seconds (926 m at equator) \citep{Fick2017}.

% Plots chosen
Plots were chosen from a larger pool of 5395 plots held in the SEOSAW database \citep{seosaw_web} based on the quality and completeness of data collection, and plot setup. Plot vegetation was identified under the broad term of `savannaj', which includes `woodland', `savanna woodland', and `tree savanna', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool. Only plots >0.1 hectares were used in analyses, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates. Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than `grassy savanna', which is considered a separate biome with very different species composition \citep{Parr2014}.

@@

% WorldClim, SoilGrids and MODIS
Climatic data were taken from the WorldClim database, using the BioClim variables \citep{Fick2017}. In addition to MAT and MAP, temperature stress was calculated as the mean diurnal temperature range (BIO2) and precipitation seasonality was calculated as the mean of the coefficient of variation of monthly mean precipitation (BIO15). Soil fertility data were extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot centre \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC) (cmolc kg\textsuperscript{-1}), soil organic carbon stocks (kg m\textsuperscript{-2}) percentage soil sand content (0.05-2 mm) by weight and soil nitrogen content (g kg\textsuperscript{-1}). These data are a modelled product derived from various remotely sensed and directly measured data sources. The degree of fire disturbance was calculated using the MODIS monthly burned area product at 500 m resolution (MCD64A1, \citealt{MODIS_burn}), counting the total number of times the plot pixel was classified as burning, between 2001 and 2018. We initially aimed to include disturbance by herbivory in our model, including total herbivore biomass from the \citet{Hempson2017} modelled herbivory product, but this inclusion prevented models from converging due to its collinearity with other observed variables, notably MAP and disturbance by fire.

@@

Estimated tree species richness was calculated for each plot using `ChaoRichness()' from the `iNEXT' package in R \citep{Hsieh2016}. This procedure extrapolates a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value. Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort among plots. Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}. To measure tree species evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} was calculated as the ratio of the estimated Shannon diversity index to the natural log of estimated species richness. Abundance evenness allows for greater niche complementarity at small scales due to potentially increased heterogeneity of functional traits. We quantified tree structural diversity for each plot by calculating the Coefficient of Variation of DBH (DBH CoV) and tree height (Height CoV).

@@

We used Structural Equation Modelling (SEM) to investigate the determinants of AGB. All SEMs were constructed and analysed in the `lavaan' package \citep{lavaan} in R version 3.6.0 \citep{R2019}. SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}. A key aspect to our decision to use SEM is that they can explicitly model and partition variance attributed to indirect effects, which is challenging in standard multiple regressions. Using SEMs also allowed us to describe latent variables such as `water availability', `soil fertility', and `disturbance' which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having directly observable measures in our dataset. SEM is also necessary to properly account for potential feedback mechanisms between aspects of environment and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to the covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.

We specified a conceptual model with factors expected to affect AGB: water availability, soil fertility, disturbance, tree species diversity, tree structural diversity and stem density (\autoref{con_mod}).

Observed variables were transformed to achieve normality where necessary and standardised to Z-scores prior to analysis (Fig. S1, Fig. S2). Standardisation allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect size, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{Beaujean2014}. Standardisation also controls for variables with variation across different orders of magnitude, which could otherwise prevent adequate model estimation from the covariance matrix in `lavaan'. To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables had their sign reversed. For example, overall water availability is expected to decrease as soil sand content increases, therefore sand content was reversed for use in the water availability latent variable. Precipitation seasonality, and temperature stress were also reversed in this way to account for the direction of their effect on water availability.

@@

We assessed the role of tree species diversity and tree structural diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity. Structural diversity can also directly influence AGB in this model, separate to the effect of of species diversity. To account for variation in stem density, which may covary with species diversity, we included it as an observed variable in our model. To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation type separately. We compared unstandardised path coefficients among the models for different vegetation types to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stem density and AGB. Path coefficients show the effect of a given path with other paths held constant. Models were estimated using the `MLM' estimator, because it is robust to multivariate non-normality \citep{Shapiro1983}. Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error of Approximation (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB. We critically assessed model fit in each case, taking into consideration the recommendations of \citet{Hu1999} who define threshold values of acceptability for these model fit indices: CFI >0.85, TLI >0.85, RMSEA <0.15, alongside our judgement of the model estimates.

@@

% Correlations
Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity and structural diversity variables had moderate positive correlations with AGB (\autoref{corr_mat}, Fig. S3). Stem density had the strongest correlation with AGB of all variables considered (\ccib{}). Environmental variables had weaker correlations with AGB than diversity variables, with all environmental variables having significant correlations with AGB, except fire frequency. The direction of these correlations was used as a test of our assumptions for the direction of influence of latent variables later used in the SEMs. MAP had positive correlations with all tree species diversity and structural diversity variables. Tree species diversity variables had clear positive correlations with stem density (species richness: \ccsi{}; Shannon equitability: \ccei{}), but structural diversity variables showed weak correlations with stem density (DBH CoV: \ccdvi{}, Height CoV: \cchvi{}).

@@

In our sub-sampling of the plot dataset by stem density, we found an increasing positive effect of tree species diversity on AGB as stem density increased (\autoref{sem_struc_stems_ha}). There appears to be a minimum stem density threshold at c. 180 trees ha\textsuperscript{-1} below which there appears to be a reasonably constant baseline effect of tree diversity on biomass. The effect of structural diversity on AGB appears to remain constant with increasing stem density. The indirect effect of tree species diversity on AGB via structural diversity climbs as stem density increases.

@@

A model incorporating the latent variables of water availability, soil fertility and disturbance by fire showed that the total effect of tree species diversity on biomass was similar to that of water availability, soil fertility and disturbance (\autoref{full_mod}, Fig. S4). The direct effects of water availability, soil fertility and disturbance on AGB were negligible (water: \fmbetamb{}, soil: \fmbetasb{}, disturbance: \fmbetafb{}), with nearly all of their observed effect on AGB coming from the indirect paths via stem density (water: \fmbetamib{}, soil: \fmbetasib{}, disturbance: \fmbetafib{}) and species diversity (water: \fmbetamd{}, soil: \fmbetasd{}, disturbance: \fmbetafd{}). MAP and soil sand content had the greatest contributions to the latent variable of water availability. Model fit was acceptable: CFI = 0.925, TLI = 0.900, and RMSEA = \fmrmsea{}, R\textsuperscript{2} of AGB = \fmrsq{}.

@@

We found a non-linear positive effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}). At low stem densities, competition between mature trees may not occur, meaning that the niche complementarity effect provided by an increase in tree species richness may not be present, accounting for the small effect of tree species diversity on AGB below c. 180 trees ha\textsuperscript{-1}. At very high stem density, there is also an increase in the effect of species diversity on structural diversity. This could be because at high stem density, the adaptation of different species to growth form become important. At low stem density, individual trees tend to spread out rather than growing tall, but at high stem density, only certain species are able to exist in the understorey, while others are able to grow tall above the woodland canopy, leading to greater variation in tree height over the plot.

@@

@@
